\section{Concept}\label{concept}

\textbf{Product type.}
The software is a distributed sensor-monitoring platform composed of autonomous peer-to-peer nodes and an optional web monitoring interface.
More precisely, the project is primarily a distributed system of executable, containerizable nodes that communicate directly through TCP sockets to exchange sensor data, membership information, and synchronization messages. The HTTP API and optional HTML dashboard are auxiliary observability surfaces layered on top of the node runtime: they are useful for inspection and monitoring, but they are not the core of the system.
Concretely, it combines:
\begin{itemize}
  \item a set of autonomous nodes that collect local sensor readings, communicate through TCP sockets, and replicate state without a central coordinator;
  \item an optional read-only web/introspection layer for monitoring and runtime observability.
\end{itemize}


\textbf{Use case description.}
The system models a smart-city scenario in which multiple nodes ingest local sensor readings, maintain a local view of shared data, and exchange updates across the peer network. Each node can be deployed close to a local sensing point, connected to one or more sensor sources, and joined to the distributed network so that locally produced readings become part of a shared, replicated monitoring view.

The platform supports two main interaction profiles:
\begin{itemize}
  \item \emph{observational interaction}: users inspect sensor values, node health, and topology evolution;
  \item \emph{operational interaction}: users configure nodes, connect sensor sources, attach nodes to the network, monitor health, and react to faults.
\end{itemize}

The system addresses three main usage profiles:
\begin{itemize}
  \item \emph{technical operators}, who administer the deployment by installing or starting node processes, configuring their network identity and bootstrap peers, connecting local sensor sources to the appropriate node, exposing the required service ports, monitoring health, and intervening when nodes fail or connectivity changes;
  \item \emph{information operators and observers}, who inspect the replicated sensor state, membership information, topology knowledge, and operational views exposed by the system in order to understand what the distributed deployment is currently reporting;
  \item \emph{automatic tools}, such as monitoring scripts, alerting components, integration services, or external dashboards, which consume the same read-only interfaces to exploit shared sensor and operational data without manually operating the distributed core.
\end{itemize}

Users interact with the system during deployment, normal operation, and fault handling. Technical operators configure nodes, sensor connections, and network reachability, while information-oriented users mainly observe the running system through read-only HTTP endpoints such as \texttt{/api/state}, \texttt{/api/membership}, and \texttt{/api/introspection}, or through the dashboard. Interaction therefore continues also during ordinary operation, even if the distributed core runs autonomously.

The intended users are trusted operators and observers working on a development machine, Docker host, or controlled laboratory network. Nodes run as local Python processes or Docker containers, while users access the exposed HTTP ports through browsers, command-line clients, automated tests, or the static dashboard. The project is therefore best understood as a reproducible distributed-systems experiment rather than as a public cloud service.

The system does not manage personal user data or application-level user records. It only handles technical runtime data such as sensor values, node metadata, membership state, metrics, and observability data.

Interaction channels and devices:
\begin{itemize}
  \item internal TCP socket communication for node-to-node coordination, replication, and membership exchange;
  \item HTTP API for programmatic monitoring and integration with external observability tools;
  \item automatic monitoring, alerting, or integration tools consuming read-only API data;
  \item desktop/laptop web browser for optional dashboard-based monitoring.
\end{itemize}

\textbf{Why distribution is needed.}
Distribution is a core requirement, not an implementation detail. Sensing points are geographically and logically distributed, so data is naturally produced at different nodes rather than at a single central site. Each node contributes local observations and shares them with peers, allowing the system to build a replicated global view without relying on a permanent coordinator.

Distribution is also needed for resource sharing and fault tolerance. Nodes share sensor state, membership knowledge, topology information, and replication metadata through peer-to-peer communication, while remaining able to operate autonomously if other peers become unreachable. This allows the system to continue local ingestion and observation during crashes or temporary partitions, and to converge again when communication is restored.

Finally, distribution is essential because the project is meant to expose real distributed-systems behavior, including delay, partial failure, recovery, and eventual consistency. These aspects are part of the project goals and must be observable rather than hidden behind a centralized architecture.

The detailed behavioral requirements, protocol constraints, and acceptance criteria are specified in \cref{requirements}.

\clearpage
