\section{Requirements Elicitation and Analysis}\label{requirements}

Requirements were elicited from the intended project goal described in \cref{concept}: build a decentralized platform for distributed sensor ingestion, replicated state propagation, fault observation, and monitoring. The elicitation process therefore focused on the observable services the platform must provide to operators and observers, on the quality attributes expected from a distributed-system prototype, and on the practical constraints imposed by course evaluation and reproducible experimentation.

\subsection{Relevant Distributed System Features}
\label{ds-features}

Not all classical distributed-system features have the same relevance in this project. \textit{Distributed Sensor Hub} is designed as a decentralized platform for collecting, replicating, and observing sensor data across multiple peer nodes, so the most important properties are decentralization, eventual convergence, fault observability, and robustness under partial failures. Other properties, such as strong security hardening or production-scale elasticity, are intentionally outside the main scope.

\textbf{Distribution transparency.}
Only partially relevant. The system should abstract low-level communication details, local sensor ingestion, and merge logic from observers and dashboard clients. However, it should not hide the fact that the system is distributed. Replication delay, membership evolution, topology knowledge, liveness suspicion, anti-entropy activity, and recovery after failures are expected to remain observable through the read-only API and dashboard. In this project, observability is more important than full distribution transparency.

\medskip
\textbf{Decentralization.}
Highly relevant. Each node should be able to run the full runtime stack independently, including local sensing, state storage, membership management, failure detection, replication, and observability. Normal operation should not rely on a master process, central coordinator, broker, or external database. Bootstrap information may be used to seed communication, but it should not introduce permanent coordinators.

\medskip
\textbf{Fault tolerance and recoverability.}
Highly relevant. Nodes may crash, restart, or become temporarily unreachable because of host failures, process restarts, or network disruption. Surviving nodes should continue ingesting local readings, serving read-only snapshots, exchanging updates with reachable peers, and maintaining their local membership view. A restarted or reconnected node should be able to recover missing state from reachable peers, using incremental synchronization when possible and a more complete state transfer when necessary. The system should prioritize continued local availability and eventual recovery over uninterrupted global consistency.

The project does not require or guarantee a fixed recovery time. Fault detection and recovery are eventual and parameter-dependent: peer suspicion depends on heartbeat timing, failure-detector thresholds, accumulated heartbeat history, and network conditions, while state recovery depends on synchronization cadence, retained update history, and the availability of at least one sufficiently up-to-date reachable peer. These behaviors should be treated as operational expectations rather than hard real-time guarantees.

\medskip
\textbf{Failure detection and fault observability.}
Highly relevant. The system should derive peer-liveness assessments from heartbeat evidence and expose states such as \texttt{alive}, \texttt{suspected}, and \texttt{dead}. These states should be visible through membership and introspection endpoints, together with recent control-plane events and replication metrics. The goal is not only to tolerate failures, but also to make failure and recovery dynamics inspectable during normal operation.

\medskip
\textbf{Consistency.}
Highly relevant. The replicated sensor state should follow an eventual consistency model. Sensor updates are disseminated asynchronously, and replicas may temporarily diverge during delays, crashes, or partitions. Convergence should be obtained through a deterministic conflict-resolution policy so that nodes receiving the same set of updates eventually expose the same winning sensor records.

This means the project does not treat transient message loss or temporary replica divergence as unacceptable in the same sense as a strongly consistent production data service would. Instead, the intended guarantee is deterministic eventual convergence once reachable nodes exchange the same update set. Because the replicated state is not required to be durable by default, recovery after extended outages may depend on at least one sufficiently up-to-date reachable peer.

\medskip
\textbf{Partition tolerance.}
Relevant. The project explicitly accepts degraded semantics during temporary communication partitions. Nodes in each connected component should continue local ingestion and observation, while replicated views may diverge. After communication is restored, synchronization mechanisms should allow reachable replicas to reconcile toward the same deterministic state.

\medskip
\textbf{Scalability.}
Relevant at the intended deployment scale. Under a full-mesh topology policy, the system is naturally better suited to small and medium deployments than to very large peer networks, because each node is expected to maintain direct connectivity with all others. The architecture should therefore preserve stable convergence, membership behavior, and observability at that scale, while remaining open to alternative injectable topology policies, such as a fixed-degree strategy, that could limit per-node connectivity and support larger networks.

\medskip
\textbf{Resource sharing.}
Highly relevant. Nodes do not share centralized hardware resources; instead, they share distributed knowledge, which is one of the core purposes of the system. This includes replicated sensor winners, peer membership information, liveness evidence, topology-related metadata, recent protocol events, and replication counters. Coordination is achieved through message exchange and deterministic merge rules rather than through a shared database or central service.

\medskip
\textbf{Openness, interoperability, and heterogeneity.}
Relevant. The system architecture should separate topology policy, sensor production, runtime processing, protocol messages, replicated state, and read-only observation clients. In particular, the connection strategy should be replaceable without forcing changes to the rest of the runtime. Sensor behavior should also remain modular, so that additional producers or adapters can be introduced while preserving the same downstream replication and observability pipeline. This makes the system open to heterogeneous sensor sources and alternative topology strategies, even though the current scope focuses on simulated local sensors rather than external industrial sensor protocols.

The system should also interoperate with external components at clear boundaries. In particular, browser-based dashboards, generic HTTP clients, automated scripts, and containerized deployment tooling should interact with the same exposed observation surfaces, while node-to-node communication remains separated behind the internal protocol.

\medskip
\textbf{Evolvability and maintainability.}
Highly relevant. The system should be organized into explicit modules for runtime orchestration, networking, protocol handling, membership, failure detection, replication, replicated state, sensors, topology, web API, and dashboard support. This separation is important because the platform combines several distributed-systems concerns at once and must remain understandable enough to validate, debug, and extend.

\medskip
\textbf{Performance, concurrency, and communication efficiency.}
Relevant, but not subject to hard real-time constraints. Each node is expected to support concurrent activities such as sensor production, state processing, heartbeat exchange, membership dissemination, replication, TCP request handling, and HTTP polling. Communication efficiency matters because excessive synchronization traffic or insufficient retained update history can degrade convergence behavior and slow recovery. Timing and synchronization parameters should therefore remain configurable. The target is stable operational behavior, not strict latency guarantees.

Accordingly, the project does not impose strict response-time or throughput requirements, nor does it define service-level guarantees. Timing behavior is intentionally parameter-dependent and should be understood as acceptable for small and medium deployments rather than as a hard real-time commitment.

\medskip
\textbf{Security and trust.}
Limited in the current scope. The platform is designed for trusted deployments under a single administrative domain, not adversarial multi-tenant environments. It does not currently provide authentication, authorization, transport encryption, Byzantine fault tolerance, or protection against malicious peers. Security hardening would be a future extension rather than a core design driver of the present implementation.

This limited scope is also consistent with the nature of the handled information: the project does not process personal, sensitive, or confidential application data, but only simulated sensor readings and technical runtime metadata such as membership state, topology views, metrics, introspection events, and logs. Multiple conceptual roles exist, such as technical operators, observers, and automated tools, but the implementation does not enforce differentiated permissions because the exposed HTTP interfaces are read-only and unauthenticated.

\medskip
\textbf{Economy and cost.}
Relevant as an implementation constraint. The system is expected to run on ordinary machines using repository-provided artifacts, Python dependencies, and Docker Compose topologies. It does not require paid cloud infrastructure, managed services, or institution-provided deployment platforms. This supports portability and low-cost deployment.

Minimizing resource usage is therefore a secondary concern rather than a primary optimization target. Bounded buffers, configurable fanout, and moderate deployment scale are useful to avoid unnecessary saturation, but the dominant objective remains dependable distributed behavior on commodity hardware.

\subsection{Glossary}
\label{glossary}

The following terms are used throughout the report with the meanings adopted by this project:
\begin{description}
  \item[\textbf{Gossip}] Best-effort dissemination of node-local knowledge, especially membership and topology information, through periodic peer-to-peer messages.
  \item[\textbf{Anti-entropy}] Periodic synchronization mechanism used to reduce divergence between replicas. In this project it is implemented as push-pull replication, where nodes both advertise local sensor deltas and request missing updates.
  \item[\textbf{Phi Accrual Failure Detector}] Adaptive heartbeat-based failure detector that converts heartbeat inter-arrival timing into a suspicion score used to classify peers as \texttt{alive}, \texttt{suspected}, or \texttt{dead}.
  \item[\textbf{Last-Writer-Wins}] Deterministic conflict-resolution policy in which the update with the greatest timestamp wins, with the origin identifier used as a deterministic tie-breaker when timestamps are equal.
  \item[\textbf{Delta}] Incremental replication entry representing a sensor-state update together with sequence, origin, timestamp, value, and metadata. Deltas are retained in a bounded in-memory buffer.
  \item[\textbf{Full sync}] Complete state transfer used when a node joins, rejoins, or can no longer recover from retained deltas alone.
  \item[\textbf{Membership}] A node-local table of known peers, including network coordinates, liveness status, heartbeat evidence, phi score, and related metadata.
  \item[\textbf{Bootstrap peer}] Initially configured peer endpoint used to seed discovery and start communication with the rest of the cluster.
  \item[\textbf{Replicated sensor state}] The node-local Last-Writer-Wins view of the latest winning sensor records known by that node.
  \item[\textbf{Introspection}] Read-only aggregate observation API exposing sensor state, membership, topology, events, and metrics using a stable schema.
  \item[\textbf{Topology}] Node-local or merged knowledge about graph-like relations among peers, used mainly for observability and connection policy.
  \item[\textbf{Eventual consistency}] Consistency model in which replicas may temporarily diverge, but converge after communication is restored and the same update set is exchanged.
  \item[\textbf{Sensor provider}] Local producer that emits sensor readings into the ingestion pipeline.
  \item[\textbf{Replication cursor}] Monotonic per-node sequence value used to request missing deltas during pull replication; it is not used for conflict resolution.
\end{description}

\subsection{Functional Requirements}
\label{functional-requirements}

\textbf{FR1 -- Multi-node autonomous operation.}
The system shall allow multiple nodes to start independently and participate in the same distributed deployment without requiring a central coordinator.

\textbf{Acceptance criterion.}
When a cluster of at least two configured nodes is started, each node becomes operational as an autonomous process and exposes its local service interfaces without depending on a single master node.

\medskip
\textbf{FR2 -- Local sensor ingestion.}
Each node shall be able to manage one or more local sensor sources or sensor connections and ingest their readings into the distributed system state.

\textbf{Acceptance criterion.}
Given at least one configured sensor on a node, the node eventually exposes fresh sensor records associated with its own origin through the read API or introspection API.

\medskip
\textbf{FR3 -- Cluster-wide sensor-state dissemination.}
The system shall propagate sensor updates among nodes through decentralized replication so that readings produced on one node become visible on other reachable nodes.

\textbf{Acceptance criterion.}
If a node generates new sensor readings while communication paths to other nodes are available, those readings eventually appear in the replicated sensor-state view exposed by the other reachable nodes.

\medskip
\textbf{FR4 -- Deterministic timestamp-based conflict resolution.}
The system shall resolve concurrent or conflicting updates through a deterministic merge policy so that replicas can converge to the same winning value for each sensor record.

\textbf{Acceptance criterion.}
When two or more updates compete for the same logical sensor record, all nodes that receive the same set of updates eventually expose the same winner for that record.

\medskip
\textbf{FR5 -- Membership discovery and peer knowledge propagation.}
The system shall allow nodes to discover peers through bootstrap and peer-to-peer membership exchange, then maintain a distributed view of cluster membership.

\textbf{Acceptance criterion.}
After startup and sufficient message exchange, a node exposes a membership view containing the peers it has discovered directly or indirectly.

\medskip
\textbf{FR6 -- Adaptive liveness observation.}
The system shall provide operators with a node-local assessment of peer liveness so that missing or degraded peers can be observed during experiments.

\textbf{Acceptance criterion.}
If a known peer becomes unreachable for long enough, at least one observing node eventually reports that peer as \texttt{suspected} or \texttt{dead} in its membership/introspection output.

\medskip
\textbf{FR7 -- Recovery after crashes or temporary disconnections.}
The system shall allow a node that restarts or reconnects after temporary unavailability to recover the replicated state needed to rejoin normal operation by synchronizing with reachable peers.

\textbf{Acceptance criterion.}
After a node crash or temporary isolation, once communication with at least one sufficiently up-to-date peer is restored, the recovering node eventually exposes a replicated state aligned with the rest of the reachable cluster.

\medskip
\textbf{FR8 -- Read-only observability interface.}
The system shall provide a read-only programmatic interface for observing sensor state, membership, topology-related information, events, and operational metrics.

\textbf{Acceptance criterion.}
A client can query documented HTTP \texttt{GET} endpoints and obtain structured snapshots of the node's current distributed-system view without modifying runtime state.

\medskip
\textbf{FR9 -- Push-pull anti-entropy exchange.}
\textbf{FR9 -- Bidirectional replication synchronization.}
The replication layer shall support bidirectional synchronization between reachable peers so that each side can both advertise local updates and obtain missing information.

\textbf{Acceptance criterion.}
During normal operation, reachable peers can send delta updates and request missing updates from one another, so each node can obtain sensor records it does not yet store.

\medskip
\textbf{FR10 -- Phi Accrual failure suspicion output.}
\textbf{FR10 -- Liveness state exposure.}
The membership subsystem shall expose liveness states allowing operators to distinguish peers reported as \texttt{alive}, \texttt{suspected}, or \texttt{dead}.

\textbf{Acceptance criterion.}
If liveness evidence from a peer degrades or stops for long enough, that peer is eventually reported as \texttt{suspected} or \texttt{dead} in membership/introspection output.

\medskip
\textbf{FR11 -- Full-state rejoin synchronization.}
\textbf{FR11 -- Rejoin state reconstruction.}
When a node rejoins after crash, restart, or long partition, it shall support recovery of a coherent replica from reachable peers even when incremental synchronization alone is insufficient.

\textbf{Acceptance criterion.}
After prolonged unavailability, a rejoining node can recover a sufficiently complete sensor-state view from a reachable peer and eventually align with the reachable cluster state, assuming such a peer still holds the relevant state.

\medskip
\textbf{FR12 -- Human-oriented monitoring support.}
The system shall support human observation of the running cluster through a dashboard or equivalent visual monitoring interface.

\textbf{Acceptance criterion.}
An observer can inspect live cluster information from a web browser through a dashboard that consumes the exposed read-only API of a node.

\subsection{Non-Functional Requirements}
\label{non-functional-requirements}

\textbf{NFR1 -- Decentralization.}
Normal operation shall not depend on any central coordinator or single control component.

\textbf{Acceptance criterion.}
The system continues to ingest local readings, maintain local state, and serve read-only observability data on surviving nodes even if one arbitrary peer becomes unavailable.

\medskip
\textbf{NFR2 -- Eventual consistency.}
The replicated sensor state shall provide eventual, not immediate, consistency across nodes.

\textbf{Acceptance criterion.}
In the absence of new conflicting updates and after communication is restored, reachable replicas converge to the same winning values for the same sensor records.

\medskip
\textbf{NFR3 -- Partition tolerance with degraded semantics.}
The system shall remain available for local operation during temporary network partitions in experimental deployments, accepting that different partitions may temporarily diverge.

\textbf{Acceptance criterion.}
During a partition, nodes in each connected component continue local ingestion and observation; after the partition heals, their replicas eventually converge again.

\medskip
\textbf{NFR4 -- Fault observability.}
The system shall make failures and recovery visible enough to support distributed-systems experimentation and debugging.

\textbf{Acceptance criterion.}
Operators can inspect status changes, replication activity, and recovery evidence through exposed membership, event, or metric views.

\medskip
\textbf{NFR5 -- Reproducibility of experiments.}
The platform shall support repeatable validation scenarios on development machines and course evaluation environments.

\textbf{Acceptance criterion.}
Using the documented setup and provided Docker Compose topologies, an operator can start a predefined cluster and reproduce representative convergence, partition, and recovery experiments without relying on external managed services.

\medskip
\textbf{NFR6 -- Extensibility of sensor sources.}
The system shall be open to additional sensor producers without changing the conceptual role of the rest of the distributed runtime.

\textbf{Acceptance criterion.}
New sensor sources can be introduced as additional producers while preserving the same downstream ingestion, replication, and observation capabilities.

\medskip
\textbf{NFR7 -- Operational scalability at course-project scale.}
The platform shall support small and medium experimental clusters rather than only a fixed two-node demo.

\textbf{Acceptance criterion.}
The documented deployment can be instantiated with multi-node topologies, such as six-node and twelve-node clusters, and still provides replicated state and observability functions for the whole experiment.

\subsection{Implementation Constraints}
\label{implementation-constraints}

\textbf{IR1 -- Self-contained local deployment.}
The project shall be executable on commodity development machines without requiring paid cloud services, institution-managed infrastructure, or external managed services.

\textbf{Justification.}
The system is intended for course-project evaluation and distributed-systems experimentation. Its deployment model must therefore remain portable and reproducible from the repository itself.

\textbf{Acceptance criterion.}
An operator can start one or more nodes using the repository code, documented dependencies, and provided deployment artifacts without provisioning external infrastructure.

\medskip
\textbf{IR2 -- Reproducible predefined topologies.}
The project shall provide predefined multi-node deployment configurations that can be reused for demonstrations, validation, and fault-injection experiments.

\textbf{Justification.}
Validation requires repeatable cluster scenarios with stable node identities, network coordinates, sensor configurations, and observation endpoints.

\textbf{Acceptance criterion.}
An operator can launch the documented predefined topologies and obtain consistent node identities, sensor sources, peer relationships, and monitoring interfaces across repeated runs.

\medskip
\textbf{IR3 -- Course-scale experimental scope.}
The implementation shall target small and medium experimental clusters suitable for local execution and course validation, rather than production-scale IoT deployments.

\textbf{Justification.}
The system is designed to demonstrate distributed-systems properties under controlled conditions, not to satisfy industrial-scale availability, performance, or device-management requirements.

\textbf{Acceptance criterion.}
The provided configurations and validation scenarios support representative multi-node deployments suitable for evaluating convergence, partition behavior, recovery, and observability at course-project scale.

\clearpage
